---
title: "River Network Distance Computation and Applications with `riverdist`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{riverdist_vignette}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8](inputenc)
---

[Package overview](#Package overview)

[Limitations, notes, and caveats](#Limitations, notes, and caveats)

[Importing and displaying river network data](#Importing and displaying river network data)

[Basic distance calculation in a non-messy river network](#Basic distance calculation in a non-messy river network)

[Incorporating flow direction](#Incorporating flow direction)

[Allowing different route-detection algorithms: a possible time-saver](#Allowing different route-detection algorithms a possible time-saver)

[Beyond individuals: summarizing or plotting at the dataset level](#Beyond individuals summarizing or plotting at the dataset level)

[Editing a river network object, or fixing a messy one](#Editing a river network object, or fixing a messy one)

[Dealing with braided channels](#Dealing with braided channels)


<a name="Package overview"></a>

Package overview
---------------------------------------
The 'riverdist' package is intended as a free and readily-available resource for distance calculation along a river network.  This package was written with fisheries research in mind, but could be applied to other fields.  The 'riverdist' package builds upon the functionality of the 'sp' and 'rgdal' packages, which provide the utility of reading GIS shapefiles into the R environment.  What 'riverdist' adds is the ability to treat a linear feature as a connected network, and to calculate travel routes and travel distances along that network.


<a name="Limitations, notes, and caveats"></a>

Limitations, notes, and caveats
--------------------------------------------
Care must be exercised in the presence of braided channels, in which multiple routes may exist between river locations.  In these cases, the default route calculation provided by 'riverdist' will be the shortest travel route, but it is possible that the shortest travel route may not be the route of interest.  Functions are provided to check for braiding, as well as select the route of interest.  If no braiding is detected, route and distance calculation can switch to a more efficient algorithm.

Another important note is that only projected data can be used, both for river networks and for point data.  The 'riverdist' environment in R does not share the ability of GIS software to project-on-the-fly, and treats all coordinates on a linear (rectangular) scale.  Therefore, the projection of all data must also be the same.


<a name="Importing and displaying river network data"></a>

Importing and displaying river network data
-------------------------------------------

[Importing a river network](#Importing a river network)

[Displaying a river network](#Displaying a river network)

[Checking connectedness](#Checking connectedness)

[Converting XY data to river locations](#Converting XY data to river locations)

[Displaying point data in river locations](#Displaying point data in river locations)



<a name="Importing a river network"></a>

**Importing a river network using `line2network()`**

In a typical workflow, a user will first import a projected polyline shapefile using `line2network()`.  This function reads the specified shapefile using the 'sp' and 'rgdal' packages, and adds the network connectivity, thus creating a river network object.  If an unprojected shapefile is detected, an error will be generated.  However, the `reproject=` argument allows the `line2network()` to re-project the shapefile before importing it as a river network.

```{r,eval=FALSE}
library(riverdist)
MyRivernetwork <- line2network(path=".", layer="MyShapefile")

# Re-projecting in Alaska Albers Equal Area projection:
AKalbers <- "+proj=aea +lat_1=55 +lat_2=65 +lat_0=50 +lon_0=-154
    +x_0=0 +y_0=0 +datum=NAD83 +units=m +no_defs +ellps=GRS80 +towgs84=0,0,0"

MyRivernetwork <- line2network(path=".", layer="MyShapefile", reproject=AKalbers)
```


<a name="Displaying a river network"></a>

**Displaying a river network using `plot()`**

Basic plotting is provided using a method of the `plot()` function.  See `help(plot.rivernetwork)` for additonal plotting arguments and graphical parameters.  Shown below is the Gulkana River network, included as a dataset.

```{r,fig.width=5,fig.height=5}
library(riverdist)
data(Gulk)
plot(x=Gulk)
```

Once the river network has been imported, some cleanup may be necessary, depending on the structure of the shapefile.  Ideally, there should be one segment between each endpoint or junction node, and the spatial extent of the shapefile should be limited to the region or network of interest.  If the user has GIS software available, it may be simplest to format the shapefile as desired before importing into R.  However, functions are included for river network formatting in R if necessary.  The `cleanup()` function interactively steps through the formatting functions in a good sequence, and may be a good way to get started.

```{r,eval=FALSE}
abstreams_fixed <- cleanup(abstreams0)
```


<a name="Checking connectedness"></a>

**Checking connectedness using `topologydots()`**

For route and distance calculation to work, the topologies must be correct, with all the right segments being treated as connected.  The `topologydots()` function can check for this, and plots connected segment endpoints as green, and non-connected endpoints as red.  This is shown below, and all appears good.

```{r,fig.width=5,fig.height=5}
topologydots(rivers=Gulk)
```


<a name="Converting XY data to river locations"></a>

**Converting XY data to river locations using `xy2segvert()` and `pointshp2segvert()`**

Once an appropriate river network has been obtained, point data can be read.  Function `xy2segvert()` converts XY data into river locations, by "snapping" each point to the closest segment and vertex.  Because of this, river locations can only be used in the context of the river network they belong to.  In addition to the river locations, the snapping distance for each observation is also returned.

The `fakefish` dataset includes coordinates from a sequence of telemetry flights.

```{r,fig.width=5,fig.height=3}
data(fakefish)
fakefish_riv <- xy2segvert(x=fakefish$x, y=fakefish$y, rivers=Gulk)
head(fakefish_riv)  # a look at the first few rows of the output
hist(fakefish_riv$snapdist, main="snapping distance (m)")
```

Spatial data can be read directly from a point shapefile using `pointshp2segvert()`, which returns the resulting river locations added to the data table from the point shapefile.

```{r,eval=FALSE}
segvert_from_shp <- pointshp2segvert(path=".", layer="MyPointShapefile", rivers=MyRivernetwork)
```


<a name="Displaying point data in river locations"></a>

**Displaying point data in river locations using `riverpoints()`**

The `riverpoints()` function works essentially like `points()` to overlay point data on an existing plot, but using river locations (segment and vertex).  The `zoomtoseg()` function produces a plot zoomed to the specified segment number, or vector of segment numbers.

In the plot below, the raw coordinates are displayed as red circles and the river locations are displayed as blue squares.

```{r,fig.width=5,fig.height=5}
zoomtoseg(seg=c(11, 14), rivers=Gulk)
points(fakefish$x, fakefish$y, pch=16, col="red")
riverpoints(seg=fakefish_riv$seg, vert=fakefish_riv$vert, rivers=Gulk, pch=15, col="blue")
```


<a name="Basic distance calculation in a non-messy river network"></a>

Basic distance calculation in a non-messy river network
-------------------------------------------------

[Computing network distance](#Computing network distance)

[Computing network distances between sequential observations of individuals](#Computing network distances between sequential observations of individuals)

[Computing a matrix of network distances between all observations of an individual](#Computing a matrix of network distances between all observations of an individual)

[Computing minimum observed home range for individuals](#Computing minimum observed home range for individuals)

[Computing a matrix of network distances between all observations](#Computing a matrix of network distances between all observations)

[Computing a matrix of network distances between all observations of two different datasets](#Computing a matrix of network distances between all observations of two different datasets)

<a name="Computing network distance"></a>

**Computing network distance using `riverdistance()`**

River network distance can be calculated directly with the `riverdistance()` function.  The `riverdistance()` function calls `detectroute()` internally, which the user will probably never need, but usage is shown below.  The `riverdistance()` function needs river locations (segment and vertex) for both starting and ending locations.  Specifying `map=TRUE` as shown below is not necessary, but can provide a check to verify that the function is working properly.

```{r,fig.width=5,fig.height=5}
# starting location: segment 7, vertex 49
# ending location: segment 14, vertex 121
detectroute(start=7, end=14, rivers=Gulk)
riverdistance(startseg=7, startvert=49, endseg=14, endvert=121, rivers=Gulk, map=TRUE)
```


<a name="Computing network distances between sequential observations of individuals"></a>

**Computing network distances between sequential observations of individuals using `riverdistanceseq()`**

River distance can be calculated manually between any two connected locations on a river network.  However, a few common summary analyses were automated and are included in 'riverdist'.  First, the `riverdistanceseq()` function returns a matrix of the distances between sequential observations for a set of individuals observed multiple times.  In the example below, three fish were observed (or not) during 5 telemetry flights.  The dataset is shown first, then the output from `riverdistanceseq()`.  Fish number 1 traveled 83.87 km between flights 3 and 4, and was not observed during flight 2.

```{r}
data(smallset)
smallset
riverdistanceseq(unique=smallset$id, survey=smallset$flight, seg=smallset$seg, 
                   vert=smallset$vert, rivers=Gulk)
```


<a name="Computing a matrix of network distances between all observations of an individual"></a>

**Computing a matrix of network distances between all observations of an individual using `riverdistancematbysurvey()`**

River distance can also be calculated between all observations of a single individual.  In the output matrix shown below for individual 1, the element with row identifier `1` and column identifier `3` represents the river distance from the location observed in survey 1 to the location observed in survey 3, calculated as 83.6 km.  An important note is that the distances reported are net distance between locations, and are not intended to be cumulative.  Specifying `full=TRUE` reports an output matrix that includes all observations, with values of `NA` if the individual was not observed.  This allows the output matrices for multiple individuals to have the same rows and columns, thus being directly comparable.


```{r}
riverdistancematbysurvey(indiv=1, unique=smallset$id, survey=smallset$flight,
      seg=smallset$seg, vert=smallset$vert, rivers=Gulk, full=FALSE)
```



<a name="Computing minimum observed home range for individuals"></a>

**Computing minimum observed home range for individuals using `homerange()`**

The minimum observed (linear) home range for each individual can be calculated using `homerange()`.  Maps can be produced if desired, by specifying `map=TRUE`.

```{r,fig.width=7,fig.height=3.5}
# calculating observed minimum home range for all individuals
par(mfrow=c(1,3))
homerange(unique=smallset$id, seg=smallset$seg, vert=smallset$vert, rivers=Gulk, map=TRUE)
```


<a name="Computing a matrix of network distances between all observations"></a>

**Computing a matrix of network distances between all observations using `riverdistancemat()`**

A matrix of the river network distance between every observation and every other observation can also be calculated using `riverdistancemat()`.  A use for this function might be if the user wishes to calculate all distances at once, and subset later.  

The `logical` argument can be used for subsetting, if the full network distance matrix is not needed.  This capability is shown below, calculating the distance matrix only for observations occurring on segment number 2.  The `ID` argument can be used with a vector of observation labels, to display row and column labels that may be easier to interpret than observation indices.

```{r}
logi1 <- (smallset$seg==2)
obsID <- paste0("id",smallset$id,"-flight",smallset$flight)  # constructing observation labels
riverdistancemat(seg=smallset$seg, vert=smallset$vert, rivers=Gulk, logical=logi1, ID=obsID)
```


<a name="Computing a matrix of network distances between all observations of two different datasets"></a>

**Computing a matrix of network distances between all observations of two different datasets using `riverdistancetofrom()`**

A matrix of river network distance between two location datasets can be calculated using `riverdistancetofrom()`, which can be similarly subsetted using the `logical1` and `logical2` arguments.  Row and column names can be added using the `ID1` and `ID2` arguments.

An example of the use of this function might be computation of distance between observations of instrumented fish and a set of fixed river locations, such as contaminant sites.  In this case, a similar function `upstreamtofrom()` (discussed later) might be used, which would give the means to calculate up- or downstream distance from contaminant sites, with the option of examining flow-connectedness.

```{r}
streamlocs.seg <- c(2,2,2)
streamlocs.vert <- c(10,20,30)
streamlocs.ID <- c("loc A","loc B","loc C")

logi2 <- (smallset$seg==2)
obsID <- paste0("id",smallset$id,"-flight",smallset$flight)

riverdistancetofrom(seg1=streamlocs.seg, vert1=streamlocs.vert, seg2=smallset$seg, 
                    vert2=smallset$vert, ID1=streamlocs.ID, ID2=obsID, logical2=logi2, 
                    rivers=Gulk)
```



<a name="Incorporating flow direction"></a>

Incorporating flow direction
-------------------------------------

[Defining flow direction of a river network](#Defining flow direction of a river network)

[Calculating flow direction and directional network distance](#Calculating flow direction and directional network distance)

[Automations of flow direction and directional network distance](#Automations of flow direction and directional network distance)

<a name="Defining flow direction of a river network"></a>

**Defining flow direction of a river network using `setmouth()`**

Flow direction and directional (upstream) distance can also be calculated.  For this to be accomplished, the segment and vertex of the river "mouth", or lowest point must first be identified.  The segment containing the river mouth can be visually identified from a plot of the river network.  In the Gulkana River example, the lowest segment happens to be segment 1.  Identifying the lowest vertex of segment 1 can be done using the plot produced by `showends()`, shown below.  In this case, the mouth vertex happens to be vertex 1.  This will not necessarily be the case.  After importing a shapefile into R, the segment vertices will be stored in sequential order, but not necessarily by flow direction.  Specifying the segment and vertex coordinates of the river network mouth can be done using `setmouth()` as shown below, though it can also be set manually by direct assignment.

```{r,fig.width=5,fig.height=5}
showends(seg=1,rivers=Gulk)
Gulk1 <- setmouth(seg=1, vert=1, rivers=Gulk)
```


<a name="Calculating flow direction and directional network distance"></a>

**Calculating flow direction using `riverdirection()` and directional network distance using `upstream()`**

If the flow direction has been established by specifying the river network mouth, river direction can be calculated using `riverdirection()`, and upstream distance can be calculated using `upstream()`.  If the input river locations are flow-connected, `riverdirection()` returns "up" if the second location is upstream of the first and "down" if downstream.  In the flow-connected case, `upstream()` returns the network distance as positive if the second location is upstream of the first, and negative if downstream.  If the input locations are not flow-connected, `riverdirection()` returns "up" if the total upstream distance is greater than the total downstream distance, and "down" otherwise.  In the non flow-connected case, upstream distance in `upstream()` can return one of two things, depending on the user's research intent.  Specifying `net=TRUE` will return the "net" distance (upstream distance - downstream distance between the two locations).  Specifying `net=FALSE` (the default) will return the total distance between the two locations, with the sign depending on whether the upstream distance exceeds the downstream distance.

For example, the route between two points goes downstream along a river for 100m, then up 20m on a tributary.  Specifying `net=TRUE` will return a distance of -80m.  Specifying `net=FALSE` will return a distance of -120m.

Specifying `flowconnected=TRUE` in both `riverdirection()` and `upstream()` will only return distances or directions if the input locations are flow-connected, and will return `NA` otherwise.

```{r,fig.width=5,fig.height=4}
zoomtoseg(seg=c(6,3), rivers=Gulk)
riverpoints(seg=c(6,4), vert=c(250,250), col=4, pch=15, rivers=Gulk1)
#riverdistance(startseg=6, endseg=4, startvert=250, endvert=250, rivers=Gulk1, map=TRUE)
text(c(859122.4, 872104.1), c(6964127.4,6969741.0), pos=c(3, 4), labels=c("beginning", "end"))
riverdirection(startseg=6, endseg=4, startvert=250, endvert=250, rivers=Gulk1)
upstream(startseg=6, endseg=4, startvert=250, endvert=250, rivers=Gulk1, net=FALSE)
upstream(startseg=6, endseg=4, startvert=250, endvert=250, rivers=Gulk1, net=TRUE)
upstream(startseg=6, endseg=4, startvert=250, endvert=250, rivers=Gulk1, flowconnected=TRUE)
```


<a name="Automations of flow direction and directional network distance"></a>

**Automations of flow direction and directional network distance**

River direction and upstream distance are also applied in functions `riverdirectionseq()` and `upstreamseq()` which work like `riverdistanceseq()`, `riverdirectionmatbysurvey()` and `upstreammatbysurvey()` which work like `riverdistancematbysurvey()`, `riverdirectionmat()` and `upstreammat()` which work like `riverdistancemat()`, and `riverdirectiontofrom` and `upstreamtofrom` which work like `riverdistancetofrom`.  The `upstreamseq()`, `upstreammatbysurvey()`, and `upstreammat()` use the additional `net=` argument, and all river direction and upstream distance functions use the additional `flowconnected=` argument.



<a name="Allowing different route-detection algorithms a possible time-saver"></a>

Allowing different route-detection algorithms: a possible time-saver
-----------------------------------------------------

Currently three algorithms are implemented to detect routes between river locations, with the functions automatically selecting the most appropriate unless another is specified.  If braiding presence is detected or unknown, Dijkstra's algorithm is used (`algorithm="Dijkstra"`) to return the shortest route, which is safe but slow.  If braiding is confirmed to be absent using `checkbraidedTF()`, the first complete route detected between locations is used, which can save considerable processing time.  

Running `buildsegroutes()` calculates the routes to each network segment from a single point, defined as the river network mouth.  This allows route calculation to switch to an even simpler method, further reducing processing time.

In the example below, distance is calculated between the two points in a very complex river network, using all three route detection algorithms, and calculating the time requirement for a single calculation.

```{r}
data(abstreams)

# distance calculation before defining braiding or building routes
tstart <- Sys.time()
riverdistance(startseg=120, startvert=10, endseg=131, endvert=10, rivers=abstreams, 
              algorithm="Dijkstra")
Sys.time()- tstart

# distance calculation before building routes
tstart <- Sys.time()
riverdistance(startseg=120, startvert=10, endseg=131, endvert=10, rivers=abstreams, 
              algorithm="sequential")
Sys.time()- tstart

# distance calculation after building routes
tstart <- Sys.time()
riverdistance(startseg=120, startvert=10, endseg=131, endvert=10, rivers=abstreams, 
              algorithm="segroutes")
Sys.time()- tstart
```

In this case, building segment routes takes about as much time as calculating a single distance, a little less than a second.  Calculating a single distance after building segment routes takes about a hundredth of a second, thus dramatically saving time if multiple distances were to be computed, such as in a large matrix of distances, or multiple analyses.


<a name="Beyond individuals summarizing or plotting at the dataset level"></a>

Beyond individuals: summarizing or plotting at the dataset level
----------------------------------------

[Calculating and plotting kernel density](#Calculating and plotting kernel density)

[Spatial data spread for each flight event](#Spatial data spread for each flight event)

[Plotting clustering or dispersal using K-functions](#Plotting clustering or dispersal using K-functions)

[Summarizing up-river position](#Summarizing up-river position)

[Plotting a summary matrix or distance sequence](#Plotting a summary matrix or distance sequence)

[Plotting all upstream movements](#Plotting all upstream movements)

<a name="Calculating and plotting kernel density"></a>

**Calculating and plotting kernel density using `riverdensity()` and `plotriverdensity()`**

A method has been provided to display kernel density calculated from point data, using river network distance.  The `makeriverdensity()` function calculates scaled kernel density at approximately regularly-spaced river network locations, with the linear resolution specified by the optional `resolution=` argument.  A gaussian (normal) kernel is used by default, but a rectangular (simple density) kernel can be used as well.  The `survey=` argument may also be used with a vector of survey identifiers corresponding to the point data.  If the `survey=` argument is used, `makeriverdensity()` will calculate separate densities for each unique survey, and a method of `plot()` will produce a separate plot for each unique survey.

The `plot()` function method can display densities using line thickness, color, or both.  For additional plotting arguments including possible color ramps, see `?plot.riverdensity`.

Densities for nine of the ten Fakefish surveys are shown below.

```{r,eval=FALSE}
data(Gulk, fakefish)
fakefish_density <- makeriverdensity(seg=fakefish$seg, vert=fakefish$vert, rivers=Gulk, 
  survey=fakefish$flight.date, resolution=2000, bw=10000)
par(mfrow=c(3,3))
plot(x=fakefish_density, ramp="blue", dark=0.85, maxlwd=15,
                 whichplots=c(1:8,10))   # showing only nine plots for clarity
```

```{r,fig.width=7.5,fig.height=8.5,echo=FALSE, eval=FALSE}
data(fakefish_density)
par(mfrow=c(3,4))
plot(x=fakefish_density, ramp="blue", dark=0.85, maxlwd=15)
```

```{r,fig.width=7.5,fig.height=10,echo=FALSE}
data(fakefish_density)
par(mfrow=c(3,3))
plot(x=fakefish_density, ramp="blue", dark=0.85, maxlwd=15,
                 whichplots=c(1:8,10))
```


<a name="Spatial data spread for each flight event"></a>

**An alternate use of `homerange()` to give the spatial data spread for each flight event**

The `homerange()` function could also be used to give a measure of the spatial spread of observations for each flight event, by using the flight identifier instead of individual identifier in the `unique=` argument.

In this case, the observations are the most closely spaced in the first flight event, and the most widely spaced in the August 11 flight event.  Interestingly, the density plot of the August 11 flight shows many of the the individuals to be located near one another, but spread to multiple tributaries, with little presence in between.  Considering this event to be the largest spread may still be appropriate, depending on the study, since all individuals were observed as concentrated in the lower mainstem on April 1.

```{r,fig.width=5,fig.height=4}
x <- homerange(unique=fakefish$flight.date, seg=fakefish$seg, vert=fakefish$vert, rivers=Gulk)
x
plot(x$range, type='b', xaxt='n', xlab="", ylab="range (m)", cex.axis=.6)
axis(1,at=1:10,labels=sort(unique(fakefish$flight.date)), cex.axis=.6, las=3)
```



<a name="Plotting clustering or dispersal using K-functions"></a>

**Plotting clustering or dispersal using K-functions, with `kfunc()`**

K-functions, defined here as the average proportion of additional individuals located within a given distance of each individual, can be a useful tool for investigating evidence of clustering or dispersal.  Calling `kfunc()` and specifying the flight indentifier will result in a sequence of plots, each displaying the K-function associated with each survey.

Unless otherwise specified, each plot will be overlayed with a confidence envelope, calculated by resampling all within-survey distances, thus creating a null distribution under the assumption that clustering is independent of survey.  Thus, a K-function above the envelope at a small distance range can be seen as evidence that locations were more clustered than expected for that survey; conversely, a K-function below the envelope at a small distance range can be seen as evidence that locations were more dispersed than expected for that survey.

The plots below show relatively high amounts of clustering for the 2015-04-01 and 2015-09-20 surveys.  The 2015-07-07 and especially 2015-08-11 surveys show possible clustering at small distances and very strong dispersal at medium distances, which can be corroborated with the density maps shown above..

```{r,fig.width=7.5,fig.height=7.5}
par(mfrow=c(3,3))
kfunc(seg=fakefish$seg, vert=fakefish$vert, rivers=Gulk, survey=fakefish$flight.date,
     maxdist=200000, whichplots=c(1:8,10))   # showing only nine plots for clarity
```



<a name="Summarizing up-river position"></a>

**Summarizing up-river position, defined as distance from mouth, using `mouthdist()` for one observation and `mouthdistbysurvey()` for a dataset**

In some cases, a meaningful summary measure may be the up-river position of each observation.  Distance between an individual observation and the river mouth can be calculated using `mouthdist()`.  In the case of multiple observations of a set of individuals, a summary matrix can be calculated using `mouthdistbysurvey()`, which returns a matrix of distances from the river mouth, with each row corresponding to a unique individual and each column corresponding to a unique survey.

```{r}
mouthdist(seg=fakefish$seg[1], vert=fakefish$vert[1], rivers=Gulk)
x <- mouthdistbysurvey(unique=fakefish$fish.id, survey=fakefish$flight,
    seg=fakefish$seg, vert=fakefish$vert, rivers=Gulk)

round(x)
```


<a name="Plotting a summary matrix or distance sequence"></a>

**Plotting the summary matrix from `mouthdistbysurvey()` or other distance sequence using `plotseq()`**

The resulting matrix of up-river position returned from `mouthdistbysurvey()`, or another distance sequence returned from `riverdistanceseq()` or `upstreamseq()`, can be plotted using `plotseq()`.  A few types of plots are available, with two shown below.  Additional `type=` arguments can be seen by calling `?plotseq`.  

The user is cautioned to use any plots returned as descriptive tools only, as ANOVA-type inference would likely be inappropriate without accounting for repeated-measures and/or serial autocorrelation.

In the example below, the plot shows evidence of the instrumented fish beginning the sampling period near the river mouth, then migrating upriver, and finally ending the sampling period near the mouth.

```{r,fig.width=7.5,fig.height=5}
x <- mouthdistbysurvey(unique=fakefish$fish.id, survey=fakefish$flight.date,
    seg=fakefish$seg, vert=fakefish$vert, rivers=Gulk)

par(mfrow=c(1,2))
plotseq(seqbysurvey=x)
plotseq(seqbysurvey=x, type="dotline")
par(mfrow=c(1,1))
```

Depending on the study, `plotseq()` may be useful in plotting the upstream distance between observations of all individual, as calculated in `upstreamseq()`.  It is worth noting that the outputs from both `upstreamseq()` and `riverdistanceseq()` may have many empty cells, if individuals were missed on any flights.  For the purpose of clarity, the default plot type is used below, which produces boxplots.  However, in this case, a jittered dotplot may be more appropriate due to the small sample sizes, and can be produced by specifying `type="dotplot"`.

In this case, the plots show evidence that movement was generally up-river between flights 1 and 2 and between flights 2 and 3, that there was the greatest variability in upstream movement between flights 5 and 6, and that movement was generally down-river in the latter part of the study.

```{r,fig.width=7.5,fig.height=5}
x <- upstreamseq(unique=fakefish$fish.id, survey=fakefish$flight, seg=fakefish$seg,
                 vert=fakefish$vert, rivers=Gulk)
round(x)
par(mfrow=c(1,2))
plotseq(seqbysurvey=x)

plotseq(seqbysurvey=x, type="dotplot")
abline(h=0, lty=3)
```


<a name="Plotting all upstream movements"></a>

**Plotting all upstream movements using `matbysurveylist()` and `plotmatbysurveylist()`**

In a more generalized application, the `matbysurveylist()` function calculates the distances or upstream distances, between each pair of observation events, for all individuals.  Functionally, this is a summary of the outputs of `riverdistancematbysurvey()` or `upstreammatbysurvey()`, but for all individuals in a dataset.  The output from `matbysurveylist()` can then be plotted using `plotmatbysurveylist()`, providing a one-plot summary of all movements within a dataset.  It is recommended to use the default `method="upstream"` in `matbysurveylist()`, which calculates directional (upstream) distances, which will result in a more informative plot.

The output plot is the upper triangle of a matrix of plots, in which plot `[i,j]` represents the upstream distances traveled between observation `i` and observation `j`, for all individuals observed in those observations.  Each plot is overlayed with a horizontal line at an upstream distance of zero, for an illustration of up-river or down-river movement trend for that pairwise movement.  It is worth noting that all sequential pairings of events (first to second, second to third, etc.) fall on the lower edge of the triangle, and this sequence of plots is the same as those given by `plotseq()` in the previous example.  

Three types of plots can be produced using the `type=` argument.  The default boxplot type (`type="boxplot"`) is shown below.  Alternately, if the `type=` argument is set to `"confint"`, each plot gives a line showing the extent of an approximate 95% confidence interval for the mean upstream distance traveled.  Inference should be made with caution, as sample sizes are likely to be small, and no attempt has been made at using a family-wise confidence level.  However, plotting confidence intervals may be further illustrative of trend.  A jittered dotplot may be produced by specifying `type="dotplot"`, which will be the most appropriate in the case of small sample sizes.  The default boxplot is shown below for the purpose of clarity, but a jittered dotplot would likely be a better choice in this instance.

```{r,fig.width=5,fig.height=10}
matbysurveylist <- matbysurveylist(unique=fakefish$fish.id, survey=fakefish$flight, seg=fakefish$seg, 
                         vert=fakefish$vert, rivers=Gulk)
plotmatbysurveylist(matbysurveylist)
```



<a name="Editing a river network object, or fixing a messy one"></a>

Editing a river network object, or fixing a messy one
----------------------------------------------------------

[All-purpose river network cleanup](#All-purpose river network cleanup)

[Issue: the river network contains unneeded or unconnected segments](#Issue the river network contains unneeded or unconnected segments)

[Issue: the river network contains sequential "runs" of segments that do not otherwise branch](#Issue the river network contains sequential runs of segments that do not otherwise branch)

[Issue: the river network segments do not break where they should](#Issue the river network segments do not break where they should)

[Issue: segments that should connect do not](#Issue segments that should connect do not)

[Issue: the river network contains segments that are smaller than the connectivity tolerance](#Issue the river network contains segments that are smaller than the connectivity tolerance)

[Issue: A long straight-line section of the river network does not contain vertices between endpoints](#Issue A long straight-line section of the river network does not contain vertices between endpoints)



Ideally, the shapefile used to define a river network object should be refined in GIS *before* importing into R.  That being said, there are sure to be instances in which it is advantageous or necessary to make changes to the river network object within R.


<a name="All-purpose river network cleanup"></a>

**All-purpose river network cleanup using `cleanup()`**

In many cases, the `cleanup()` function will be the recommended first step in fixing a messy river network object after importing it.  The `cleanup()` function should be called within the console, and interactively calls the editing functions in sequence.  It then returns a new, edited river network object which can be edited further, or used as is.  It may even be the most straightforward to call `cleanup()` multiple times.

Usage will look like

```{r,eval=FALSE}
data(abstreams0)  # a messy river network
abstreams_fixed <- cleanup(abstreams0)  # fixing many problems
```


<a name="Issue the river network contains unneeded or unconnected segments"></a>

**Issue: the river network contains unneeded or unconnected segments - fixes using `trimriver()`, `trimtopoints()`, and `removeunconnected()`**

This is a very likely issue, particularly if the network was imported without changes in a GIS environment.  River network segments can be manually removed using `trimriver()`, as shown below.  Using the argument `trim=` removes the specified segments, and using the argument `trimto=` removes all *but* the specified segments.  The example below is fairly simplistic, but illustrates the usage of the `trim=` and `trimto=` arguments.

```{r,fig.width=7.5,fig.height=4, warning=FALSE}
Gulk_trim1 <- trimriver(trim=c(10,11,12,13,14), rivers=Gulk)
Gulk_trim2 <- trimriver(trimto=c(10,11,12,13,14), rivers=Gulk)

par(mfrow=c(1,3))
plot(x=Gulk, main="original")
plot(x=Gulk_trim1, main="trim=c(10,11,12,13,14)")
plot(x=Gulk_trim2, main="trimto=c(10,11,12,13,14)")
```

It is also possible to trim a river network to include only segments that are within a spatial tolerance of a set of X-Y points.  The `trimtopoints()` function offers three methods of doing this.  Specifying `method="snap"` (the default) returns a river network made up only of the closest segments to the input points.  This is the simplest method, but may result in spatial gaps, as shown in the example below.  Specifying `method="snaproute"` returns a network of the closest segments to the input points, but also includes any segments necessary to maintain a connected network.  Specifying `method="buffer"` returns a river network made up of segments with endpoints or midpoints located within a specified "buffer"  distance of the input points.  This may be advantageous if the user wants to include segments that are near, but not directly proximal, to the input points.

```{r,fig.width=5,fig.height=5}
data(Kenai3)
x <- c(174185, 172304, 173803, 176013)
y <- c(1173471, 1173345, 1163638, 1164801)
Kenai3.buf1 <- trimtopoints(x=x, y=y, rivers=Kenai3, method="snap")
Kenai3.buf2 <- trimtopoints(x=x, y=y, rivers=Kenai3, method="snaproute")
Kenai3.buf3 <- trimtopoints(x=x, y=y, rivers=Kenai3, method="buffer", dist=5000)

plot(x=Kenai3, main="original")
points(x, y, pch=15, col=4)
legend(par("usr")[1], par("usr")[4], legend="points to buffer around", pch=15, col=4, cex=.6)
```

```{r,fig.width=7.5,fig.height=3.5}

par(mfrow=c(1,3))
plot(x=Kenai3.buf1, main="snap")
points(x, y, pch=15, col=4)
plot(x=Kenai3.buf2, main="snaproute")
points(x, y, pch=15, col=4)
plot(x=Kenai3.buf3, main="buffer, dist=5000")
points(x, y, pch=15, col=4)
```

Automatically removing all segments not connected to the river network mouth can be done using `removeunconnected()`.  This may be useful if extraneous lines are retained from an import from GIS.  This function may take some time to process, and simplifying the network using `dissolve()` is recommended.

```{r,fig.width=7.5,fig.height=5}
data(Koyukuk2)
Koy_subset <- trimriver(trimto=c(30,28,29,3,19,27,4),rivers=Koyukuk2)
Koy_subset <- setmouth(seg=1,vert=427,rivers=Koy_subset)

Koy_subset_trim <- removeunconnected(Koy_subset)

par(mfrow=c(1,2))
plot(x=Koy_subset, main="original")
plot(x=Koy_subset_trim, main="unconnected segments removed")
```


<a name="Issue the river network contains sequential runs of segments that do not otherwise branch"></a>

**Issue: the river network contains sequential "runs" of segments that do not otherwise branch - a fix using `dissolve()`**

Using an unnecessarily complex river network can greatly increase processing time.  Runs of segments can be combined using `dissolve()`, which works much like a spatial dissolve within GIS.

```{r,fig.width=7.5,fig.height=5, warning=FALSE}
data(Kenai2)
Kenai2_sub <- trimriver(trimto=c(26,157,141,69,3,160,2,35,102,18,64,86,49,103,61
                                 ,43,183,72,47,176), rivers=Kenai2)

Kenai2_sub_dissolve <- dissolve(rivers=Kenai2_sub)

par(mfrow=c(1,2))
plot(x=Kenai2_sub, main="original")
plot(x=Kenai2_sub_dissolve, main="dissolved")
```


<a name="Issue the river network segments do not break where they should"></a>

**Issue: the river network segments do not break where they should - a fix using `splitsegments()`**

This issue is problematic, as it directly affects how connectivity is detected within the river network.  Without appropriate connectivity, routes and distances cannot be calculated.  To address this issue, `splitsegments()` automatically breaks segments where another segment endpoint is detected.

In the example below, segments 7, 8, 13, and 16 need to be split in multiple places.  Since connectedness is not detected for the associated tributaries, `topologydots()` shows the endpoints as red, or unconnected.  Calling `splitsegments()` breaks the segments in the appropriate places, allowing for the network to be connected as it should be.

```{r,fig.width=7.5,fig.height=4.5, warning=FALSE}
data(Koyukuk1)

Koyukuk1.split <- splitsegments(rivers=Koyukuk1)

par(mfrow=c(1,2))
topologydots(rivers=Koyukuk1, main="original")
topologydots(rivers=Koyukuk1.split, main="split")
```


<a name="Issue segments that should connect do not"></a>

**Issue: segments that should connect do not - a fix using `connectsegs()`**

Segments can be manually "attached" at their closest points using `connectsegs()`.

```{r,fig.width=7.5,fig.height=4, warning=FALSE}
data(Koyukuk0)

Koyukuk0.1 <- connectsegs(connect=21, connectto=20, rivers=Koyukuk0)

par(mfrow=c(1,2))
plot(Koyukuk0, ylim=c(1930500,1931500), xlim=c(194900,195100), main="original")
topologydots(Koyukuk0, add=TRUE)

plot(Koyukuk0.1,ylim=c(1930500,1931500), xlim=c(194900,195100), main="connected")
topologydots(Koyukuk0.1, add=TRUE)
```


<a name="Issue the river network contains segments that are smaller than the connectivity tolerance"></a>

**Issue: the river network contains segments that are smaller than the connectivity tolerance - a fix using `removemicrosegs()`**

This is an issue that may be difficult to recognize, and may cause mysterious problems with network topology, sometimes preventing route calculation.  These "microsegments" can be removed using `removemicrosegs()`.

```{r,eval=FALSE}
data(abstreams1)
abstreams2 <- removemicrosegs(abstreams1)
```


<a name="Issue A long straight-line section of the river network does not contain vertices between endpoints"></a>

**Issue: A long straight-line section of the river network does not contain vertices between endpoints - a fix using `addverts()`**

In some cases, such as when a river network contains a lake, the shapefile will contain long straight-line sections, with vertices retained only for the beginning and end.  If point data exist in these regions, when they are converted to river locations, they will be snapped to the nearest vertex - in this case, one of the endpoints of the straight stretch.  Since it is likely that a greater degree of precision is desired in distance calculations, `addverts()` provides a method of inserting additional vertices, wherever the river network contains vertices that are spaced at a greater distance than a specified threshold.

The example below first shows how the Skilak Lake section of the Kenai River network was originally read from a shapefile, with the vertices of segment 71 overlayed. The second plot shows the same section with the vertices of segment 71 overlayed, after adding vertices every 200 meters to the full river network.

```{r,fig.width=7.5,fig.height=4, warning=FALSE}
data(Kenai3)
Kenai3split <- addverts(Kenai3, mindist=200)

par(mfrow=c(1,2))
zoomtoseg(seg=c(73,70,71), rivers=Kenai3, main="segment 71 vertices")
points(Kenai3$lines[[71]]) 

zoomtoseg(seg=c(73,70,71), rivers=Kenai3split, main="adding points every 200m")
points(Kenai3split$lines[[71]])  
```


<a name="Dealing with braided channels"></a>

Dealing with braided channels
--------------------------------
Most of the utility of the 'riverdist' package was designed assuming a truly dendritic river network, in which there is no braiding of channels and only one path exists between one river location and another.  If this is not the case and a braided network is used, the user is **strongly cautioned** that the distances reported may be inaccurate.  In the event of braiding, the shortest route between two locations is returned, but the possibility exists that this may not be the route desired.

[Checking for braiding](#Checking for braiding)

[Investigating multiple routes](#Investigating multiple routes)

<a name="Checking for braiding"></a>

**Checking for braiding using `checkbraided()`**

Braiding can be checked for in a river network as a whole using `checkbraided()`, which can take a while to run on a large or complex network.  In the example below, no braiding exists in the Gulkana River network, and severe braiding exists in the Killey River West channel network.

```{r,fig.width=7.5,fig.height=5}
data(Gulk, KilleyW)
par(mfrow=c(1,2))
plot(x=Gulk, main="Gulkana River")
plot(x=KilleyW, ylim=c(1164500, 1168500), main="Killey River West")

checkbraided(rivers=Gulk)
checkbraided(rivers=KilleyW)
```

Braiding can also be checked for specific routes.  In the example below, braiding does not exist between segments 1 and 7, but does exist between segments 1 and 5.

```{r,fig.width=5,fig.height=5}
Kenai3.subset <- trimriver(trimto=c(18,1,64,27,104,93,91,83,45,2), rivers=Kenai3)
plot(x=Kenai3.subset)

checkbraided(startseg=1, endseg=7, rivers=Kenai3.subset)
checkbraided(startseg=1, endseg=5, rivers=Kenai3.subset)
```

<a name="Investigating multiple routes"></a>

**Investigating multiple routes using `riverdistancelist()`**

If the user wishes to explore the possibility of multiple routes beween two locations, a non-robust function `routelist()` detects a list of routes from one segment to another by means of randomization, which is applied by function `riverdistancelist()` to calculate the distance along the routes that were detected.  Because randomization is used to generate possible routes, the list of routes it generates is not guaranteed to be complete.  The `reps=` argument can be used to specify the number of randomizations to run, with more reps resulting in a more complete list of routes, but taking more processing time.  The `riverdistancelist()` function, shown below, returns a list of all routes detected by ascending distance, and the corresponding distances.

```{r}
Killey.dists <- riverdistancelist(startseg=1,endseg=16,startvert=25,endvert=25,
   rivers=KilleyW,reps=1000)
Killey.dists  # 18 routes are detected.
```

The shortest and longest routes detected are mapped below.

```{r,fig.width=7.5,fig.height=5}
par(mfrow=c(1,2))
plot(x=KilleyW, ylim=c(1164500, 1168500), main="shortest route")
riverdistance(startvert=25, endvert=25, path=Killey.dists$routes[[1]], 
              rivers=KilleyW, map=TRUE, add=TRUE)
plot(KilleyW, ylim=c(1164500, 1168500), main="longest route")
riverdistance(startvert=25, endvert=25, path=Killey.dists$routes[[18]], 
              rivers=KilleyW, map=TRUE, add=TRUE)
```

It is worth noting that the default functions for route and distance calculation do return the shortest route, in the presence of multiple possible routes.  The default-calculated route is shown below, and is the same as the shortest route determined by `riverdistancelist()`.

```{r}
detectroute(start=1, end=16, rivers=KilleyW)
Killey.dists$routes[[1]]  #calculated above
```